---
title: "Introduction to musicMCT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to musicMCT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(musicMCT)
```

# Introduction

The goal of `musicMCT` is to provide computational tools for the study
of musical scales. It includes functions that will probably be familiar to
music theorists from pitch-class set theory in the vein of Allen Forte's
*The Structure of Atonal Music* (1973). For instance, we can calculate the 
**i**nterval-class **vec**tor of the diatonic scale (Forte's **s**et **c**lass 7-35)
as follows:

```{r}
ivec(sc(7,35))
```

The top line of the block above is what you would input to R; the bottom line is the
output of the function. Ignore the `[1]`, which just indicates that you're
seeing the beginning of the output. The ic-vector is 2 5 4 3 6 1 as we'd expect.

The main purpose of the package, however, is not to reproduce traditional pc-set theory
but to let us explore the geometry of musical scales described in "Modal Color Theory" 
(Sherrill 2025, *Journal of Music Theory* 69/1: 1-49). As that article explains, 
seven-note scales live in a 6-dimensional geometry populated by 1,824,229 qualitatively 
distinct scalar structures. I'm happy to calculate an `ivec()` by hand, but we're
going to need computers to explore such a complicated space!

This vignette will introduce many of the main functions of `musicMCT` that serve
this purpose, especially as they relate to the concerns of the *Journal of Music Theory* article. The
problem that we'll discuss is not of deep significance, but I hope a toy example
will teach you how to use this software and give you a sense of the kinds of
questions that deeper study might get into.


# How acoustic is the "acoustic scale"?
## 1. Introducing the main characters

A familiar object from the theory of 20th-century music is the scale (C, D, E,
F-sharp, G, A, B-flat). This is the fourth mode of melodic minor. We can verify
that fact by defining the melodic minor scale for R and then calling `sim()`,
the package's function for finding the modes of a scale:

```{r}
melodic_minor <- c(0, 2, 3, 5, 7, 9, 11)
sim(melodic_minor)
```

Note the top line's syntax, which defined the melodic minor scale: we'll be using it a lot! Pitch-classes
are represented by numbers: C=0, C-sharp = 1, and so on. For familiar sets in 12edo
(12 **e**qual **d**ivisions of the **o**ctave), these are all integers. Later
we'll see that it's easy to work with integers in any k edo. But the values that define
our scales don't have to be integers at all: Modal Color Theory works in continuous 
pitch-class space.

The function `sim()` above computes the **s**calar **i**nterval **m**atrix of a set 
(as defined by [Tymoczko 2008](https://doi.org/10.1111/j.1468-2249.2008.00257.x)), which 
presents to us the modes of a scale as the columns of the matrix. Our input, the melodic
minor scale, is in the first column, and the scale we want to study can be read from the fourth
column of the SIM:

```{r}
sim(melodic_minor)[,4]
```

Just to flex our muscles in R, let's pretend that we've forgotten how to interpret pitch-class integers
and that we therefore need to verify that those numbers do correspond to the 
notes (C, D, E, F-sharp, G, A, B-flat). We'll do that by checking the **v**oice **l**eading from
C major to this scale:

```{r}
c_major <- c(0, 2, 4, 5, 7, 9, 11)
minimizeVL(c_major, sim(melodic_minor)[,4])
```

As [Tymoczko 2007](https://www.mtosmt.org/issues/mto.05.11.4/mto.05.11.4.tymoczko.pdf) points
out, key signatures are essentially just voice leadings from the C major scale to another heptachord,
and here we've found a voice leading that raises the fourth step (F) a semitone while lowering the seventh step (B).
So the numbers (0, 2, 4, 6, 7, 9, 10) do indeed correspond to the scale (C, D, E, F-sharp, G, A, B-flat).

Now, this scale is sometimes called the "acoustic" scale because in some sense it is
close to a seven-note chunk of the overtone series. Our task in this vignette is to use `musicMCT`
to see how real that similarity is. Let's agree to call these two scales the "acoustic" and "overtone" scales, respectively.

The overtone scale is defined by treating the 7th through 13th harmonics of the overtone series as a scale
(noting that the 14th harmonic is an octave above the 7th). We'll convert that into semitone measurements
as follows:

```{r}
overtones <- 7:13
frequency_ratios <- overtones / 7
semitone_values <- 12 * log2(frequency_ratios)
overtone_scale <- sim(semitone_values)[,2]
print(overtone_scale)
```

Let's find the voice leading from the overtone scale to the acoustic scale, since voice leading distance
offers an approximate measure of musical similarity (as [Callender, Quinn, and 
Tymoczko 2008](https://doi.org/10.1126/science.1153021) argue):

```{r}
acoustic_scale <- sim(melodic_minor)[,4]
minimizeVL(overtone_scale, acoustic_scale)
```

Intuitions about voice leading distances can be tricky, but this initially doesn't seem like a 
big distance between the two scales. The average amount that each voice has to move is about 23
cents (a quarter of a semitone). On the other hand, our voice leading from C major to
C acoustic was also pretty small: the average distance an individual voice had to move there was
about 29 cents (2/7 of a semitone). Thus, by one measure, the `acoustic_scale` is about as good of
an approximation of `c_major` as it is of the `overtone_scale`!

We need better tools than mere voice-leading distance to think about scale similarity. One is the
idea of quantization, which we'll address in the next section. We'll see that it doesn't tell
the whole story either.

## 2. A few supporting characters

The normal justification for calling the `acoustic_scale` "acoustic" is that it's what you
get if you simply round the values of the `overtone_scale` to the nearest integer (in 12edo):

```{r}
round(overtone_scale, digits=0)
```

Wait a second, that's not even the acoustic scale! It has two consecutive semitones (6, 7, 8) and
actually represents sc7-33, the "whole-tone plus one" scale. What's going on here!?

[Tymoczko (2013, 130)](https://doi.org/10.1080/17459737.2013.818724) explains that there's really 
no such thing as *the* quantization of the `overtone_scale`. There are 7 different quantizations. The
rounding above unfairly privileges C (`0`) as a note that's guaranteed not to move. If we treat 
all the degrees of the scale more equally, sometimes a different note gets to stay fixed. Let's try it.

If the scale's tonic starts on a 12edo integer, it won't change when we quantize the scale. Transposing
the whole scale up 1 semitone by using `tn()`, we should expect to get essentially the same quantization, just with 1 added to every value: 

```{r}
round(overtone_scale, digits=0)
round(tn(overtone_scale, 1), digits=0)
```

To find the other possible quantizations, we need to explore the range of transpositions where the scale's
tonic is some value ``0 < x < 1``. If we transpose the `overtone_scale` so that it starts on every cent (i.e.
hundredth of a semitone) in that range, that should be a fine enough sampling to encounter every quantization:

```{r}
amounts_to_transpose <- (0:99)/100
transposed_scales <- sapply(amounts_to_transpose, tn, set=overtone_scale)
quantized_scales <- apply(transposed_scales, 2, round, digits=0)
unique_quantizations <- unique(quantized_scales, MARGIN=2)
print(unique_quantizations)
```

We're almost there, but some of the scales start on 0 and some on 1. The eighth column is just ${T_1}$ of 
the first. So let's make them all start on zero to double-check that they're truly unique:

```{r}
unique_quantizations_from_0 <- apply(unique_quantizations, 2, startzero)
final_quantizations <- unique(unique_quantizations_from_0, MARGIN=2)
colnames(final_quantizations) <- apply(final_quantizations, 2, fortenum)
print(final_quantizations)
```

And thus we really do get exactly the seven quantizations that Tymoczko predicts. Of these seven, is there any reason
to think that the `acoustic_scale` is a better quantization than the other possibilities? Tymoczko proposes one
perspective: of the 100 scales in `transposed_scales`, how *many* of them quantize to each of the seven `final_quantizations`?
This is easy to answer too (though here I'm going to gloss over the details of the code that gets these numbers):

```{r, echo=FALSE}
quantization_labels <- apply(final_quantizations, 2, fortenum)
quantization_weights <- diff(which(duplicated(quantized_scales, MARGIN=2)==FALSE))
quantization_weights[1] <- quantization_weights[1] + 1
names(quantization_weights) <- quantization_labels
print(quantization_weights)
```

That is, the WT-plus-1 scale that we first quantized to accounts for about 11% of the range of quantizations, whereas there
are two distinct modes of melodic minor (sc7-34) which together account for 54% of the quantizations. The first mode, which is
the `acoustic_scale` proper, accounts on its own for 37%--definitely the lion's share of them.

This lends credence to conventional view, suggesting that the `acoustic_scale` might be a preferable
quantization of the `overtone_scale` even if it isn't the *only* quantization.


## 3. But what about scale structure?

So far we've used some of the computational tools of `musicMCT` but we haven't really applied the concepts
of Modal Color Theory. That's because we've been treating the scales' continuous voice-leading space as 
essentially undifferentiated, whereas the central argument of MCT is that there are discrete regions in the geometry
that correspond to *qualitative* differences in scale structure.

MCT models scale structure by comparing intervals that belong to the same generic size. For instance, a big part of the
character of the familiar major scale lies in the fact that most of its steps have the same size, except for $\hat{3}$-$\hat{4}$
and $\hat{7}$-$\hat{1}$ which are smaller (but match each other). MCT breaks this down into individual comparisons: is the step
$\hat{1}$-$\hat{2}$ bigger or smaller than the step $\hat{2}$-$\hat{3}$? Is the skip $\hat{1}$-$\hat{3}$ bigger or smaller than
the skip $\hat{3}$-$\hat{5}$? And so on.

How long before we have exhausted all the potential comparisons that we could make? Each comparison corresponds to a hyperplane
in the geometry, and `musicMCT` offers a complete list of all the hyperplanes in a matrix called an `ineqmat` (for "inequality matrix").
The relevant matrix for four-note scales looks like this:

```{r}
getineqmat(4)
```

Each row of the matrix represents a different pairwise interval comparison: for tetrachords, we apparently need to consider 8
different comparisons to have a complete accounting of a scale structure. Two scales have the same structure (or belong to the same
"color") if they answer all 8 comparisons in the same way. `musicMCT` summarizes this information in a scale's `signvector()`. For
instance, let's consider set classes 4-6 (prime form 0127) and 4-24 (prime form 0248). Here are their sign vectors:

```{r}
signvector(sc(4,6))
signvector(sc(4,24))
```

These vectors are the same, so MCT considers the tetrachords to have the same scalar structure. For instance, both start with two identical steps
(0-1-2 and 0-2-4) and then have a larger leap (2-7 and 4-8).

For heptachords, the relevant `ineqmat` has 42 rows, so the space of scale structures is considerably more complicated. Generally 
you wouldn't learn a lot from trying to read sign vectors directly--`musicMCT` has more human-readable functions that will help you
interpret them--but let's take a gander at them for the acoustic and overtone scales:

```{r}
signvector(acoustic_scale)
signvector(overtone_scale)
```

One visually apparent difference between these is that the `acoustic_scale` has many more values of `0` in its sign vector than the `overtone_scale` 
does. In fact, the latter has only a *single* `0`: all the rest of its values are `1` or `-1`. What does this tell us about
the structure of the two scales?

Each entry in a sign vector is a comparison between two intervals in a scale: both `1` and `-1` mean that one interval is bigger than
the other, whereas a `0` means that the two intervals are identical. For instance, the first `0` in the sign vector for 0127 and 0248
is what encodes the fact that both of them have a first step that equals their second step.

From this, we can see that the `acoustic_scale` is considerably more *regular* than the `overtone_scale`, in the sense that it has 
many specific interval sizes that repeat inside it (like the 5 whole tones that dominate its generic steps, and the 4 perfect fifths
that are the majority of its generic fifths). This contrast between the scales isn't too surprising. The `acoustic_scale` is constrained
to twelve-tone equal temperament, so there are only so many intervals that it could possibly be made up of. By contrast, the `overtone_scale`
is defined in continuous pitch-class space and has a lot more variety to choose from. Moreover, given the logarithmic relationship between
frequency ratios and pitch intervals, each successive step in the overtone series is smaller than the previous, so it makes sense
that the `overtone_scale` would have a lot of intervallic variety. If anything, it's surprising that it repeats *any* intervals.

We can learn what interval it does repeat by putting together the information from the sign vector with the `ineqmat` for heptachords. We could
manually count which entry of the sign vector is a 0, but `musicMCT` also has a function for that:

```{r}
whichsvzeroes(overtone_scale)
```

We can now look at the row of the heptachordal `ineqmat` to see what interval comparison it defines:

```{r}
getineqmat(7)[whichsvzeroes(overtone_scale),]
```

This tells us that the fourth above $\hat{5}$ equals the fourth below $\hat{5}$ in size. (See page 43 of "Modal Color Theory" for a discussion
of how to read the rows of an `ineqmat` like this.) And, sure enough, this is true:

```{r}
signed_interval_class(overtone_scale[5]-overtone_scale[1])
signed_interval_class(overtone_scale[5]-overtone_scale[2])
```

If the `acoustic_scale` is to be an approximation of the `overtone_scale`, we might want it to retain this repeated interval, which it does:
the perfect fourth from D to G matches the perfect fourth from G to C. We could also verify that by checking that the 38th entry of its sign vector
is also 0:

```{r}
signvector(acoustic_scale)[38]
```

In fact, why don't we consider all seven of the distinct quantizations from the previous section?

```{r}
all_signvectors <- apply(final_quantizations, 2, signvector)
all_signvectors[38,]
```

Most of them do match the `overtone_scale` on this point, except for sc7-31 and sc7-28. Note that these are also the scales that
barely occurred at all as quantizations: each accounts for only about 2% of all the quantizations according to the table at the
end of the previous section.

We can do better than looking at *only* position `38` in the sign vector. Why don't we compare the entire sign vectors at once? There
are various ways to do this, but one that's simple to calculate is to treat the sign vectors as if they were real vectors in a 
42-dimensional space and compute their distances:

```{r}
signvectors_with_overtone_scale <- rbind(signvector(overtone_scale), t(all_signvectors))
rownames(signvectors_with_overtone_scale)[1] <- "o.s."
dist(signvectors_with_overtone_scale)
```

The first column above, labeled "o.s." for `overtone_scale`, represents the distance between the `overtone_scale`'s sign vector
and the 7 other sign vectors (as labeled on the rows of the matrix). The other columns represent distances among the various
quantizations.

In a surprise upset, the quantization with the closest sign vector to the `overtone_scale` is apparently sc7-27 as represented by
its mode (0, 2, 4, 5, 7, 8, 9). This doesn't mean that sc7-27 is a *better* quantization than the `acoustic_scale`, just that it's
one that better reflects the structure of the `overtone_scale`. (In particular, it comes closest to capturing the fact that
the `overtone_scale`'s step sizes get smaller as you go up the scale. We'll return to this consideration as we approach our
final approximation at the end of the next section.) The modes of melodic minor simply have too much regularity--
too many zeroes in their sign vectors--to be very structurally similar to the `overtone_scale`.

Another scalar property that we might consider is how many degrees of freedom a scale has ("Modal Color Theory," 26-27): how freely can
you vary its individual scale degrees while maintaining the same overall structure? A minimally constrained heptachord has six degrees of 
freedom, and we know that the `overtone_scale` has one zero in its sign vector, which ought to make it lose a degree of freedom. (Our choices
about $\hat{1}$ and $\hat{5}$ ought to determine where $\hat{2}$ goes.) Let's compare the `overtone_scale`'s freedom to its seven quantizations:

```{r}
howfree(overtone_scale)
apply(final_quantizations, 2, howfree)
```

All of these are much less free to vary than the `overtone_scale`! All but sc7-31 can only vary along their line of saturation ("Modal Color Theory,"
20), and sc7-31 is only a little bit more flexible. This is a consequence of quantizing to twelve-tone equal temperament, which simply doesn't have
enough distinct notes to accommodate more flexible heptachords. We can check the degrees of freedom of all 38 heptachordal set classes in 12edo:

```{r}
all_12edo_heptachords <- sapply(1:38, sc, card=7)
heptachord_freedoms <- apply(all_12edo_heptachords, 2, howfree)
table(heptachord_freedoms)
```

This table shows that there are 11 set classes with 1 degree of freedom and 27 set classes with 2 degrees of freedom. None are freer than that.
If we were looking for an approximation of the `overtone_scale` with similar structural properties (like 5 degrees of freedom), we unwittingly
set ourselves up to fail by restricting our attention to the twelve-tone chromatic universe.


## 4. Looking for better approximations outside of 12edo

If we are willing to quantize to other equal divisions of the octave, we can find much better structural approximations. In particular, `musicMCT`
has a function that will find a quantized representative of any scalar color, in the lowest n-note equal division of the octave where such a
quantization exists.

```{r}
quantized_overtone_color <- quantize_color(overtone_scale)
print(quantized_overtone_color)
```

Note that this function returns a `list` object. The first entry of the list defines the scale itself: pitch class integers 0, 6, 11, 15, 18, 20, 21.
The second entry tells us interpret those integers in 30 equal divisions of the octave. If we want to use this as an input to other functions, usually
we only want the pitch-class values themselves, which we can access as `quantized_overtone_color$set`. We'll need to tell most functions to interpret
that information in 30edo separately, usually by setting the function's `edo` parameter explicitly. 

We should check this quantization in two ways. Let's convert it to twelve-tone equal temperament to get a sense for how objectively close it is
to the `overtone_scale`, and let's verify that it belongs to the same "color" as the `overtone_scale` by verifying that it has the same sign vector
as the scale it approximates:

```{r}
round(overtone_scale, digits=2)
convert(quantized_overtone_color$set, 30, 12)
isTRUE(all.equal(signvector(overtone_scale), signvector(quantized_overtone_color$set, edo=30)))
```

So this quantization does indeed have all the same structural properties as the `overtone_scale`, but its actual pitches place it rather farther 
away than the `acoustic_scale` was. This just reinforces the point that there's no such thing as one single quantization of a given scale: there
are only different quantizations that satisfy different equivalences we might choose to care about.

In that spirit, there's yet another way to approximate the `overtone_scale` which we haven't yet considered. This form of approximation uses
the notion of *adjacency* between colors in the hyperplane arrangement ("Modal Color Theory," 31-34). If two scalar colors are adjacent, one
is a simplification of the other that preserves some aspects of scalar structure while also enforcing a greater degree of regularity on the
scale. For instance, the major triad (0, 4, 7) and the minor triad (0, 3, 7) are non-equivalent structures that are both adjacent to the "neutral"
triad (0, 3.5, 7). The neutral triad has only two sizes of step (3.5 and 5) as opposed to the three step sizes for major and minor (3, 4, and 5).
It has a similar structure to them, in that its last step (5 semitones from fifth to root) is larger than the others, but it is more regular
than major and minor because it makes no distinction between its root-third and third-fifth intervals. Another color adjacency which may be familiar
to music theorists is the relationship between the 5-limit just diatonic scale (Ptolemy's "intense diatonic") and the meantone diatonic scale represented
by (0, 2, 4, 5, 7, 9, 11) in 12edo. The just and meantone scales are similar in many ways, but the meantone scale elides distinctions that the just
diatonic makes, such as the difference between major and minor whole tones. (The meantone diatonic is *named* for this particular elision of difference,
since it averages out the sizes of the larger and smaller just whole tones.)

With this in mind, let us ask: what more regular colors is the `overtone_scale` adjacent to? We might first compare it to the 12edo quantizations 
above. `musicMCT` lets us do this with the function `comparesignvecs()`, which returns `1` if two sign vectors represent adjacent colors, `0` if 
two sign vectors are identical, and `-1` if the colors are neither the same nor adjacent.

```{r}
overtone_sv <- signvector(overtone_scale)
apply(all_signvectors, 2, comparesignvecs, signvecY=overtone_sv)
```

Alas, none of our 12edo quantizations are adjacent colors to the `overtone_scale`! This isn't surprising given the table of sign-vector distances
near the end of the previous section, but it reinforces our conclusion that twelve-tone equal temperament isn't the temperament you'd choose
if your only goal were to approximate these 7 overtones.

How do we find colors that *are* adjacent to the `overtone_scale`? It isn't as simple as making a minimal change to the sign vector and creating a
scale to match, since most conceivable sign vectors don't correspond to scales that can actually exist. We could try to solve this problem
from scratch, but luckily we don't have to: part of MCT is a map of all the colors that exist (for scale sizes up to heptachords) and their 
adjacency relationships.

This map involves datasets that are too large to distribute with the `musicMCT` package. They can be downloaded from the [GitHub repository
modalcolortheory](https://github.com/satbq/modalcolortheory):

* [representative_scales.rds](https://github.com/satbq/modalcolortheory/blob/main/representative_scales.rds) (~80 MB) contains a list of all scalar colors for scales of 2-7 notes, with one concrete scale to represent each color
* [representative_signvectors.rds](https://github.com/satbq/modalcolortheory/blob/main/representative_signvectors.rds) (~10 MB) contains a list of their corresponding sign vectors
* [color_adjacencies.rds](https://github.com/satbq/modalcolortheory/blob/main/color_adjacencies.rds) (~130 MB) contains information about the adjacency relationships between all these colors (in the form of an adjacency list)

If you wish to work with this data, you should save the files in your working directory, which you can find by calling `getwd()` in R. Then import
the files into your R session with the following commands:

```{r, eval=FALSE}
representative_scales <- readRDS("representative_scales.rds")
representative_signvectors <- readRDS("representative_signvectords.rds")
color_adjacencies <- readRDS("color_adjacencies.rds")
```

The file `representative_signvectors` is particularly useful because it enables us to assign a well-defined "color number" to identify each
color in a MCT hyperplane arrangement ("Modal Color Theory," 28). If you try to look up a scale's color number without the `representative_signvectors`
saved, you'll probably get a `NULL` result:

```{r}
colornum(overtone_scale)
```

but if you do have the file, you'll discover that the `overtone_scale` belongs to heptachord color `1824025`. From this, we can look up an accounting
of all colors that are adjacent to it. `color_adjacencies` is structured as a *list* in R, whose top level organization reflects the size of the
scales you are studying, so `color_adjacencies[[7]]` contains all the information about the geometry of seven-note scales. To find out which scales are
adjacent to color `1824025`, we want the list element `color_adjacencies[[7]][[1824025]]`. The following code won't work on your device unless you
have `color_adjacencies` saved and loaded into R, but its objective is to find the most regular scale that the `overtone_scale` is structurally adjacent
to:

```{r, eval=exists("color_adjacencies") && exists("representative_scales") && exists("representative_signvectors")}	
os_adjacent_colors <- color_adjacencies[[7]][[colornum(overtone_scale)]]
os_adjacent_scales <- representative_scales[[7]][, os_adjacent_colors]
regularity <- apply(os_adjacent_scales, 2, countsvzeroes)
most_regular_approximation <- os_adjacent_scales[, which.max(regularity)]
quantize_color(most_regular_approximation)
```

If that code runs on your device, it should tell you that the most regular quantized approximation of the `overtone_scale` is the scale 
(0, 2, 4, 6, 8, 10, 11) in 14-tone equal temperament. Converted to familiar semitone measurements, this is the scale (0.00, 1.71, 3.43, 5.14,
 6.86, 8.57, 9.43). To understand better how this scale "regularizes" the `overtone_scale`, let's consider the two scales in terms of their
ranked step sizes:

```{r}
asword(overtone_scale)
best_simple_approximation <- convert(c(0, 2, 4, 6, 8, 10, 11), 14, 12)
asword(best_simple_approximation)
```

This tells us that the `overtone_scale` starts with its 6th largest step and then ascends through progressively smaller step sizes, until
its final step from $\hat{7}$ to $\hat{1}$ uses the scale's largest step size.

Our `best_simple_approximation` reduces most of that variety out of the step sizes: all the steps from its $\hat{1}$ up to $\hat{6}$ are equal.
The only intervallic variety that this approximation retains is that $\hat{6}$ to $\hat{7}$ is still the smallest step in the scale, and
$\hat{7}$ to $\hat{1}$ is still the largest step in the scale.

By the logic of color adjacency, this is the most regular scale that simplifies the structure of the `overtone_scale` without contradicting
any of its structure. By contrast, our original `acoustic_scale` does flatly contradict aspects of the `overtone_scale`'s structure. For instance,
in the `acoustic_scale` the step from $\hat{4}$ to $\hat{5}$ is smaller than the step from $\hat{5}$ to $\hat{6}$, but in the `overtone_scale` $\hat{5}$
to $\hat{6}$ is smaller (because all its step sizes get smaller until the last one). We can see that contradiction by comparing the 
10th place of their respective sign vectors:

```{r}
signvector(overtone_scale)[10]
signvector(best_simple_approximation)[10]
signvector(acoustic_scale)[10]
```

The `acoustic_scale` is the odd one out here. It's also, in terms of voice-leading distance, not clearly closer to the `overtone_scale`
than our `best_simple_approximation` is:


```{r}
round(minimizeVL(overtone_scale, acoustic_scale), 3)
round(minimizeVL(overtone_scale, best_simple_approximation), 3)
```

It's not easy to eyeball which one of those is a *smaller* overall voice leading, partially because the answer to that question depends on what
measure of distance you want to use. (For those who care about such things, the `acoustic_scale` is a closer approximation using the taxicab norm,
but our `best_simple_approximation` is closer under the Euclidean and Chebyshev norms.)

## 5. A tempered conclusion

It would make for a tighter narrative to end with a conclusive declaration that our `best_simple_approximation` really is the best way to quantize
the `overtone_scale`. But reality isn't so simple! The best approximation for a scale depends on what you're trying
to approximate for. This needs to be underscored, because the attempt to justify certain scale structures as mathematically ideal
is probably the oldest tradition of European music theory. By using math to talk about scales, Modal Color Theory may look like it contributes
to that tradition, when my goal with the theory is to do just the opposite. The questions that MCT should be useful for are descriptive ones like
"If you're using this scale, what can you do with it?" and "If you want to do X, how can we find some scale Y to facilitate X?" [Sami Abu Shumays](https://mtosmt.org/issues/mto.24.30.4/mto.24.30.4.shumays.html) is right to decry as a fallacy "the idea that the tunings of musical scales are the result of mathematical laws, rather than cultural choices that intersect with acoustical realities." My hope for MCT is that its math lets us get
deeper insight into the nature of that intersection between cultural choice and acoustical affordance.

In the end, the `acoustic_scale` actually is a pretty decent approximation for many purposes, especially if you're limited to the familiar 12 pitch-classes
of the piano. Hopefully, though, this vignette shows how we might, when musicking with different priorities, come to make other choices--and how `musicMCT` can aid us in that process.

# Coda: Brightness graphs

Up to this point, we've done everything numerically, but "Modal Color Theory" also has some visualization tools. The one I want to introduce here is the "brightness graph," 
which lets us visualize the voice-leading relationships between a scale's modes (and thus to some degree the internal structure of the scale, as described in "Modal Color 
Theory," 7-11). Using `musicMCT` we can make quick mock-ups of brightness graphs with `brightnessgraph()`. First let's take a look at the `acoustic_scale`:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
brightnessgraph(acoustic_scale)
```

The information in this graph is displayed in the same way as Figure 2b of "Modal Color Theory" (p. 8), though you'll notice
a few differences if you compare the article's figure to the one in this vignette. One difference is trivial: the mode at the bottom
of our figure here is mode IV whereas it's listed as mode VII in the *Journal of Music Theory* article. That's just because the article treats melodic minor
as mode I whereas we've asked for the modes of the `acoustic_scale`. (If you replace `acoustic_scale` with `melodic_minor` in the command
above, you should get an identical graph with the roman numeral labels shifted around.) Either way, the precise pitch values for
all the nodes are the same: the bottom mode in both graphs is (0, 1, 3, 4, 6, 8, 10).

The less trivial difference between the article and graph vignettes is their overall shape. The function `brightnessgraph()` determines
the vertical position of graph nodes exactly as "Modal Color Theory" describes, but it has less guidance for how to make decisions about
left-right placement of nodes. It produces graphs that are usually legible but which may not clarify the underlying structure: I tend
to use the function as a first draft and then construct my own (ideally more revealing) graphs by hand. For instance, the Figure 2b
of "Modal Color Theory" represents all $T_2$ relationships as vertical arrows, resulting in a graph whose overall shape is a trapezoid.
`brightnessgraph()` obscures the predominance of $T_2$ in the structure of the scale by making the chain of whole steps bow out to the left
at the middle, giving the graph a diamond-like shape.

Let's now take a look at the brightness graph for the `overtone_scale`:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
brightnessgraph(overtone_scale, numdigits=1, show_sums=FALSE)
```

This looks pretty messy! The graph is less connected (i.e. there are fewer arrows) and the vertical placement of its
nodes seems haphazard. This reflects the fact that the overtone scale has less regularity as measured by its degrees of freedom
or the number of zeroes in its `signvector()`. Nevertheless, the graph still has some useful information. Since the height of a
mode corresponds to its sum brightness, we can see those comparisons at a glance. We can also use sum brightness relationships
as a proxy for the relative height of individual pitches in the scale: if a pitch serves as the tonic of a very bright mode,
the pitch itself is very low ("dark") in relation to the scale. Thus, since mode VII is at the top of our graph here, $\hat{7}$ 
must be the darkest (relatively lowest) pitch of the scale. Conversely, since mode III is at the bottom of the graph, $\hat{3}$
is the brightest individual note of the scale.

Recall that one of the approximations we tested above was a quantization of the `overtone_scale` to a division of the octave
into 30 steps. Let's see what the brightness graph of that approximation looks like:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
brightnessgraph(quantized_overtone_color$set, edo=30, show_sums=FALSE)
```

That's impressively close to the graph for the actual `overtone_scale`! (In truth, we got a little lucky here: many colors offer
enough room for variation that `brightnessgraph()` doesn't always preserve such visually obvious similarity.) Aside from
a check that `quantize_color()` works as intended, one other thing to note from this example is that `brightnessgraph()`, like
almost all functions in `musicMCT` makes it convenient to work in other "chromatic cardinalities" (i.e. number of unit steps to
the octave) besides 12edo. Just be careful not to make direct comparisons between absolute values reckoned in different systems:
note, for instance, that the sum brightnesses for our 30edo quantization are all around 90 whereas they were about 36 in the 
brightness graph of the original `overtone_scale`. That's because we've changed our unit of measurement to be 2.5 times smaller,
so values like distances and sums scale correspondingly.

You might find it instructive to explore the brightness graphs of the remaining `final_quantizations` from above, which show
considerable variety from one to the next. This gives a glimpse of the range of possible structures for seven note scales. Simply
change the number `4` in the line of code below to any value from `1` to `7` to check the different scales:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
brightnessgraph(final_quantizations[,4])
```

Finally, let's take a look at the brightness graph for our `best_simple_approximation`. I'm going to suppress its listing of the 
pitch-classes in each mode, since without manual tweaking they overlap to the point of illegibility:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
brightnessgraph(best_simple_approximation, show_pitches=FALSE, show_sums=FALSE)
```

This feels right to me. At a glance, it's a lot like the graph for the `overtone_scale` but without all the mess. All the modes beneath VII
are now neatly lined up on a single row at the bottom. That's the visual equivalent of our search for regularity and color adjacency in the 
previous section. This is something I like about brightness graphs: they summarize, in a picture or two, ideas that one could spend several
thousand words developing.