---
title: "Bright notes, dark scales, and musicMCT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bright notes, dark scales, and musicMCT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(musicMCT)
```

This vignette is meant to be an entry point to musicMCT for readers of the article "Dark notes, bright scales, and 
Debussy's *La fille aux cheveux de lin*." It explains the tools in the package that are useful in applying the
ideas of the article, and it recreates some of the specific calculations there.

# Main Concepts

Throughout the article and this vignette, the familiar major scale will be a useful test case:

```{r}
c_maj <- c(0, 2, 4, 5, 7, 9, 11)
```

Note R's syntax for entering an ordered tuple, `c()`, which is always necessary when manually defining a musical set.

Another important scale is the perfectly even division of the octave, represented in the article as $\mathcal{E}_n$. This can be
found using the function `edoo()`:

```{r}
augmented_triad <- edoo(3)
diminished_seventh <- edoo(4)
equiheptatonic <- edoo(7)

augmented_triad
diminished_seventh
equiheptatonic
```

The main concept of the article is the function $\Phi_n$, which calculates the brightness of all a scale's notes. This function
is implemented in musicMCT by `phi()`. Simply enter the scale you want to calculate between the parentheses of `phi()`. (The parameter
$n$ of $\Phi_n$ doesn't need to be specified, since it's obvious from the number of notes in your scale.)

Compare the following code block to the article's Example 2.4:

```{r}
phi(c(0, 4, 7))
phi(c(2, 6, 9))
phi(c_maj)
```

$\Phi_n$ also goes by another name in musicMCT, `coord_to_edo()`, which emphasizes the function's sense as a change of coordinate 
system. Its inverse, going from a list of brightness values to a musical scale (i.e., a tuple in [cpitch]{.smallcaps}$^n$), is 
`coord_from_edo()`.

We can find the darkest note of a scale by a built-in R command:

```{r}
which.min(phi(c_maj))
```

This tells us that $\hat{4}$, the subdominant, is the darkest note of the major scale. Like conventional music theory, R is a 1-indexed
language, so it considers $\hat{1}$ to be the tonic and $\hat{4}$ to be the subdominant. (In the article, which uses 0-indexing, the 
subdominant is $\hat{3}$ and the minimal brightness value is $\varphi_3$.) Also, be aware that R's `which.min()` function returns only
the *first* minimum value, which may obscure the existence of brightness ties:

```{r}
minor_seventh <- c(0, 3, 7, 10)
phi(minor_seventh)
which.min(phi(minor_seventh))
```

Finally, the **sum brightness** $\beta$ of a scale can be found with the function `sum_brightness()`. By default, this gives the brightness of
the mode you entered, but the `mode` parameter can specify a different mode (0-indexed, following Definition 3.1). We can test this 
following the examples in Section 3.

Example 3.8:
```{r}
major_triad <- c(0, 4, 7)
sum_brightness(major_triad)
sum_brightness(major_triad, mode=1) # first inversion
sum_brightness(major_triad, mode=2) # second inversion
```
 
Example 3.9:
```{r}
sum_brightness(c_maj)
sum_brightness(c_maj, mode=3) # lydian
sum_brightness(c_maj, mode=5) # aeolian
```

Example 3.10:
```{r}
H <- c(0, 1, 3, 4, 6, 7, 8, 10)
phi(H)
sum_brightness(H)
sum_brightness(H, mode=7)
subset_varieties(c(0, 2, 4, 6), H, unique=FALSE)
```

The last command of the code block above calculates all the seventh chords that are embedded in $H$ as a stack of thirds.
These are represented as columns: the first column shows the quality of seventh chord built on the first step of $H$, 
$(0, 3, 6, 8)$; the second column shows the chord quality on the next step of the scale, $(0, 3, 6, 9)$; and so on. Note how
the dominant seventh chord $(0, 4, 7, 10)$ only occurs once, over the scale's penultimate note ($A\flat$).

A useful tool for visualizing scalar brightness is the **brightness graph**. This visual representation isn't discussed in
"Dark notes," but it is introduced on pp. 7–11 of ["Modal Color Theory"](https://doi.org/10.1215/00222909-11595194). In a 
brightness graph, the vertical positioning of a mode corresponds to its sum brightness. Thus, brightness ties can be 
seen from a brightness graph when multiple nodes in the graph are on a horizontal line. This lets us visualize the many brightness
ties of Example 3.11, for instance:

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
#| fig.alt: >
#|   A brightness graph for the Guidonian hexachord.
#|   Three equally bright modes (the fourth, fifth, and sixth) are
#|   on a horizontal line at the top of the figure. Three equally dark
#|   modes (the first, second, and third) are on a line at the bottom.
brightnessgraph(c(0, 2, 4, 5, 7, 9))
```

So far, all of our examples have been in twelve-tone equal temperament. Scales in other tunings are equally easy to study,
though, and musicMCT offers quick access to the 5-limit just diatonic scale of Example 3.16 by the command `j(dia)`.

```{r}
phi(j(dia))

alpha <- function(heptachord) heptachord - c_maj
del <- c(-2, 0, 2, -3, -1, 1, 3)
spc_numbers <- function(heptachord) 7*alpha(heptachord) + del

round(spc_numbers(j(dia)), 2)
```

# Hyperplane Arrangements

Section 3.3 discusses how note brightness (and sum brightness) can be characterized in terms of hyperplane arrangements,
like I do for voice-leading brightness in "Modal Color Theory." In musicMCT, the "braid" arrangement $\mathcal{B}_n$ that 
describes note brightness is called the "white" hyperplane arrangement. The augmented matrix defining the hyperplanes themselves
can be created with `make_white_ineqmat()`:

```{r}
make_white_ineqmat(3)
```

However, we often want to work with more human-readable functions like `signvector()` and `howfree()`. In this case, you can
specify that musicMCT should use the white arrangement (rather than the MCT arrangement) with the `ineqmat` parameter of many functions.

For instance, consider the trichords described in the last paragraph of Section 3.3. Their similarities and differences, reckoned against
two different hyperplane arrangements, can be seen here:

```{r}
A <- c(0, 2, 5)
B <- c(0, 1, 6)
B_prime <- c(0, 1, 7)

phi(A)
phi(B)
phi(B_prime)

signvector(A, ineqmat="white")
signvector(B, ineqmat="white")
signvector(B_prime, ineqmat="white")

signvector(A) # or signvector(A, ineqmat="mct") if we want to specify the MCT arrangement explicitly
signvector(B)
signvector(B_prime)
```

# Debussy's *La fille aux cheveux de lin*

The package musicMCT is designed to compute properties of scales, not to automate the process of music analysis. There is therefore
relatively little from the final section of the paper that can be recreated in this vignette. The main thing that can be reproduced
here is the information in Table 1:

```{r}
trichord_M <- c(0, 4, 7)
tetrachord_I <- c(0, 4, 7, 9)
pentachord_P <- c(0, 2, 4, 7, 9)
pentachord_P_prime <- c(0, 2, 5, 7, 9)
hexachord_G <- c(0, 2, 4, 5, 7, 9)
heptachord_D <- c(0, 2, 4, 5, 7, 9, 11)

phi(trichord_M)
phi(tetrachord_I)
phi(pentachord_P)
phi(pentachord_P_prime)
phi(hexachord_G)
phi(heptachord_D)
```
